---
title: "Introduction to bayestrials"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to bayestrials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The `bayestrials` package provides a comprehensive framework for Bayesian reanalysis of clinical trials. This vignette introduces the basic workflow and demonstrates key features of the package.

```{r setup}
library(bayestrials)
```

## Basic Workflow

### 1. Model Specification

One of the key features of `bayestrials` is the ability to specify models using CSV files. Here's an example model specification:

```{r}
# Read model specifications from CSV files
model_specs <- read_model_spec(
  system.file("extdata", "model_specs.csv", package = "bayestrials"),
  system.file("extdata", "prior_specs.csv", package = "bayestrials"),
  system.file("extdata", "covariate_specs.csv", package = "bayestrials")
)

# Look at the primary model
model_specs$primary_model
```

You can also create model specifications programmatically:

```{r}
# Create a model specification
my_model <- create_model_spec(
  outcome = "response",
  predictors = "treatment + age + sex",
  random_effects = "(1|center)",
  family = "bernoulli",
  link = "logit",
  model_name = "my_custom_model"
)

# Specify priors
my_priors <- PriorSpecification()
my_priors <- add_prior(my_priors, "b_treatment", "normal", 0, 1)
my_priors <- add_prior(my_priors, "b_age", "normal", 0, 2)
my_priors <- add_prior(my_priors, "sd_center", "half_cauchy", NA, 5, 0)

# Attach priors to model
attr(my_model, "priors") <- my_priors
```

### 2. Data Preparation

`bayestrials` includes functions for loading and preprocessing clinical trial data:

```{r}
# Load simulated data (in a real scenario, you would use load_clinical_trial)
set.seed(42)
n <- 200
data <- data.frame(
  id = 1:n,
  treatment = factor(sample(c("Control", "Treatment"), n, replace = TRUE)),
  age = rnorm(n, 50, 10),
  sex = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  center = factor(sample(c("Site A", "Site B", "Site C"), n, replace = TRUE)),
  response = rbinom(n, 1, 0.5)
)

# Preprocess data
preprocessed_data <- preprocess_data(
  data,
  system.file("extdata", "covariate_specs.csv", package = "bayestrials")
)
```

For multi-trial analyses, you can harmonize variables across trials:

```{r}
# Create simulated trial data
trial1 <- data.frame(
  id = 1:100,
  arm = factor(sample(c("Control", "Treatment"), 100, replace = TRUE)),
  age_years = rnorm(100, 50, 10),
  gender = factor(sample(c("Male", "Female"), 100, replace = TRUE)),
  site = factor(sample(c("A", "B", "C"), 100, replace = TRUE)),
  outcome = rbinom(100, 1, 0.5)
)

trial2 <- data.frame(
  id = 1:120,
  trt = factor(sample(c("Control", "Treatment"), 120, replace = TRUE)),
  age = rnorm(120, 55, 12),
  sex = factor(sample(c("M", "F"), 120, replace = TRUE)),
  center = factor(sample(c("1", "2", "3"), 120, replace = TRUE)),
  resp = rbinom(120, 1, 0.6)
)

# Harmonize trials
trials <- list(trial1 = trial1, trial2 = trial2)
harmonized_trials <- harmonize_trials(
  trials,
  mapping_file = system.file("extdata", "variable_mapping.csv", package = "bayestrials")
)
```

### 3. Fitting Models

Once you have prepared your data and specified your model, you can fit it using `fit_model()`:

```{r}
# Fit model
model <- fit_model(model_specs$primary_model, preprocessed_data)

# Check model
model
```

For sensitivity analyses, you can fit multiple models with different specifications:

```{r}
# Define sensitivity specifications
sensitivity_specs <- list(
  model_specs$primary_model,
  model_specs$sensitivity_1
)

# Run sensitivity analyses
sensitivity_results <- run_sensitivity_analyses(
  model,
  preprocessed_data,
  sensitivity_specs,
  type = "model"
)
```

### 4. Analyzing Results

`bayestrials` provides functions for extracting and visualizing model results:

```{r}
# Extract posterior summaries
posterior_summary(model)

# Plot posterior distribution
plot_posterior(model, variables = "b_treatment")

# Check model diagnostics
diagnostics <- check_diagnostics(model)
```

### 5. Identifying Effect Modifiers

You can test for treatment effect heterogeneity:

```{r}
# Identify effect modifiers
modifiers <- identify_effect_modifiers(
  model,
  preprocessed_data,
  treatment_var = "treatment",
  candidate_modifiers = c("age", "sex")
)

# Plot interaction
plot_interaction(modifiers, "age")
```

### 6. Generating Reports

`bayestrials` can generate comprehensive reports for models:

```{r}
# Generate basic report
report_file <- generate_report(model, template = "basic_report")

# Generate sensitivity report
sensitivity_report_file <- sensitivity_report(sensitivity_results)
```

## Conclusion

This vignette provided a brief introduction to the `bayestrials` package. For more detailed examples and advanced features, please refer to the other vignettes and function documentation.

## Session Info

```{r}
sessionInfo()
```