---
title: "Introduction to bayestrials"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to bayestrials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The `bayestrials` package provides a comprehensive framework for Bayesian analysis of clinical trials using the [brms](https://paul-buerkner.github.io/brms/) backend. This vignette introduces the basic workflow and demonstrates key features with working examples.

```{r setup}
library(bayestrials)
```

## Basic Workflow

### 1. Generate Example Data

Let's start with generating some synthetic clinical trial data for demonstration:

```{r generate-data}
set.seed(42)
data <- generate_synthetic_data(
  n_observations = 100,  # Small dataset for quick compilation
  outcome_type = "continuous",
  seed = 42
)

# Look at the data structure
head(data)
str(data)
```

### 2. Model Specification

Create a model specification programmatically:

```{r model-spec}
# Create a simple linear regression model
model_spec <- create_model_spec(
  outcome = "outcome_continuous",
  predictors = "treatment + age_centered + baseline_severity_centered",
  family = "gaussian",
  model_name = "simple_trial_model"
)

print(model_spec)
```

### 3. Prior Specification

Specify informative priors using the `PriorSpecification` system:

```{r priors}
# Create priors
priors <- PriorSpecification()
priors <- add_prior(priors, "b_treatment", "normal", 0, 2.5)
priors <- add_prior(priors, "b_age_centered", "normal", 0, 1)
priors <- add_prior(priors, "b_baseline_severity_centered", "normal", 0, 2)
priors <- add_prior(priors, "Intercept", "normal", 0, 10)
priors <- add_prior(priors, "sigma", "half_cauchy", NA, 5, 0)

print(priors)

# See how priors are converted for brms
brms_priors <- to_brms_prior(priors)
print(brms_priors)

# Attach priors to model
attr(model_spec, "priors") <- priors
```

### 2. Data Preparation

`bayestrials` includes functions for loading and preprocessing clinical trial data:

```{r}
# Load simulated data (in a real scenario, you would use load_clinical_trial)
set.seed(42)
n <- 200
data <- data.frame(
  id = 1:n,
  treatment = factor(sample(c("Control", "Treatment"), n, replace = TRUE)),
  age = rnorm(n, 50, 10),
  sex = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  center = factor(sample(c("Site A", "Site B", "Site C"), n, replace = TRUE)),
  response = rbinom(n, 1, 0.5)
)

# Preprocess data
preprocessed_data <- preprocess_data(
  data,
  system.file("extdata", "covariate_specs.csv", package = "bayestrials")
)
```

For multi-trial analyses, you can harmonize variables across trials:

```{r}
# Create simulated trial data
trial1 <- data.frame(
  id = 1:100,
  arm = factor(sample(c("Control", "Treatment"), 100, replace = TRUE)),
  age_years = rnorm(100, 50, 10),
  gender = factor(sample(c("Male", "Female"), 100, replace = TRUE)),
  site = factor(sample(c("A", "B", "C"), 100, replace = TRUE)),
  outcome = rbinom(100, 1, 0.5)
)

trial2 <- data.frame(
  id = 1:120,
  trt = factor(sample(c("Control", "Treatment"), 120, replace = TRUE)),
  age = rnorm(120, 55, 12),
  sex = factor(sample(c("M", "F"), 120, replace = TRUE)),
  center = factor(sample(c("1", "2", "3"), 120, replace = TRUE)),
  resp = rbinom(120, 1, 0.6)
)

# Harmonize trials
trials <- list(trial1 = trial1, trial2 = trial2)
harmonized_trials <- harmonize_trials(
  trials,
  mapping_file = system.file("extdata", "variable_mapping.csv", package = "bayestrials")
)
```

### 4. Fitting the Bayesian Model

Now let's fit the model using the brms backend:

```{r fit-model}
# Fit the model (using minimal settings for speed)
model <- fit_model(
  model_spec = model_spec,
  data = data,
  chains = 1,      # Single chain for speed
  iter = 500,      # Fewer iterations for demo
  cores = 1        # Single core for reproducibility
)

# Check the fitted model
print(model)
```

### 5. Model Diagnostics

Check convergence and model diagnostics:

```{r diagnostics}
# Check diagnostics
diagnostics <- check_diagnostics(model)
print(diagnostics)

# Extract estimates
estimates <- extract_estimates(model)
print(estimates)
```

### 6. Visualizing Results

Create visualizations of the posterior distributions and results:

```{r visualizations}
# Plot posterior distribution for treatment effect
plot_posterior(model, variables = c("b_treatment"))

# Create forest plot of estimates  
forest_plot <- plot_forest(estimates)
print(forest_plot)
```

## Key Features of bayestrials

### Prior Conversion System

bayestrials automatically handles prior conversion for brms compatibility:

```{r prior-conversion}
# half_cauchy priors are automatically converted to student_t
original_prior <- PriorSpecification()
original_prior <- add_prior(original_prior, "sigma", "half_cauchy", NA, 5, 0)

converted <- to_brms_prior(original_prior)
print(converted)
```

### Built-in Diagnostics

The package provides comprehensive model checking:

```{r model-summary}
# Get posterior summary
summary_stats <- posterior_summary(model)
head(summary_stats)
```

## Next Steps

This vignette covered the basic workflow for fitting a single Bayesian model. For more advanced features, explore:

- **Sensitivity Analysis**: Comparing models with different prior specifications
- **Multi-trial Analysis**: Harmonizing and analyzing multiple clinical trials
- **Effect Modifiers**: Identifying treatment effect heterogeneity
- **Custom Priors**: Advanced prior specification techniques

## Session Info

```{r session-info}
sessionInfo()
```