---
title: "Getting Started with bayestrials"
author: "bayestrials Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bayestrials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to FALSE initially to avoid compilation issues during package building
)
```

# Getting Started with bayestrials

This guide will help you install and set up bayestrials, then walk through your first Bayesian clinical trial analysis.

## Installation

### System Requirements

Before installing bayestrials, ensure you have the necessary system components:

#### R Version
bayestrials requires R version 4.0.0 or higher:

```{r check-r-version}
# Check your R version
R.version.string
```

#### C++ Compiler for Stan

bayestrials uses Stan (via brms) which requires a C++ compiler:

**macOS:**
```bash
# Install Xcode Command Line Tools
xcode-select --install
```

**Windows:**
- Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) for your R version
- Ensure "Add rtools to system PATH" is checked during installation

**Linux (Ubuntu/Debian):**
```bash
sudo apt update
sudo apt install build-essential
```

### Package Installation

```{r install-bayestrials}
# Install development version from GitHub
if (!require("devtools")) install.packages("devtools")
devtools::install_github("AartikSarma/bayestrials")

# Load the package
library(bayestrials)
```

### Verify Installation

Test that everything is working:

```{r verify-installation}
# Check package version
packageVersion("bayestrials")

# Test basic functionality
data <- generate_synthetic_data(n_observations = 10, seed = 123)
head(data)
```

## Your First Analysis

Let's walk through a complete analysis from start to finish.

### Step 1: Generate or Load Data

```{r first-analysis-data}
# Generate synthetic clinical trial data
set.seed(42)
trial_data <- generate_synthetic_data(
  n_observations = 150,
  outcome_type = "continuous", 
  seed = 42
)

# Examine the data
str(trial_data)
summary(trial_data[c("treatment", "age_centered", "outcome_continuous")])
```

### Step 2: Specify Your Model

```{r first-analysis-model}
# Create a simple linear regression model
model_spec <- create_model_spec(
  outcome = "outcome_continuous",
  predictors = "treatment + age_centered",
  family = "gaussian",
  model_name = "first_model"
)

print(model_spec)
```

### Step 3: Set Up Priors

```{r first-analysis-priors}
# Create informative but reasonable priors
priors <- PriorSpecification()

# Treatment effect: expecting small to moderate effect
priors <- add_prior(priors, "b_treatment", "normal", 0, 2.5)

# Age effect: expecting small effect
priors <- add_prior(priors, "b_age_centered", "normal", 0, 1.0)

# Intercept: centered around expected outcome value
priors <- add_prior(priors, "Intercept", "normal", 10, 5)

# Residual standard deviation: half-Cauchy (converted to student_t)
priors <- add_prior(priors, "sigma", "half_cauchy", NA, 5, 0)

print(priors)

# Attach priors to model specification
attr(model_spec, "priors") <- priors
```

### Step 4: Fit the Model

```{r first-analysis-fit}
# Fit the Bayesian model
# Using conservative settings for first run
result <- fit_model(
  model_spec = model_spec,
  data = trial_data,
  chains = 2,      # Start with 2 chains
  iter = 1000,     # 500 warmup + 500 sampling per chain
  cores = 1        # Use 1 core initially
)

# Print model summary
print(result)
```

### Step 5: Check Diagnostics

```{r first-analysis-diagnostics}
# Check convergence and diagnostics
diagnostics <- check_diagnostics(result)
print(diagnostics)

# Extract parameter estimates
estimates <- extract_estimates(result)
print(estimates)
```

### Step 6: Visualize Results

```{r first-analysis-plots, fig.width=7, fig.height=5}
# Plot posterior distributions
plot_posterior(result, variables = c("b_treatment", "b_age_centered"))

# Create forest plot
forest_plot <- plot_forest(estimates)
print(forest_plot)
```

## Understanding Your Results

### Interpreting Estimates

The `extract_estimates()` function returns:

- **Estimate**: Posterior mean
- **Est.Error**: Posterior standard deviation  
- **Q2.5**, **Q97.5**: 95% credible interval bounds
- **Rhat**: Convergence diagnostic (should be ≈ 1.00)
- **Bulk_ESS**, **Tail_ESS**: Effective sample sizes

### Key Diagnostics to Check

1. **Rhat**: Should be < 1.01 for all parameters
2. **ESS**: Should be > 400 for reliable estimates
3. **Divergences**: Should be 0 (or very few)
4. **Energy**: BFMI should be > 0.2

## Common Issues and Solutions

### Installation Problems

**Problem**: Stan compilation fails
**Solution**: 
```{r troubleshoot-compilation}
# Check compiler tools
Sys.which("make")
Sys.which("g++")  # Linux/Mac
Sys.which("gcc")  # Windows with Rtools

# If empty, install/reinstall compiler tools
```

**Problem**: Package installation fails
**Solution**:
```{r troubleshoot-install}
# Update R and try again
install.packages(c("devtools", "brms", "rstan"))
devtools::install_github("AartikSarma/bayestrials")
```

### Model Fitting Issues

**Problem**: Model won't converge (high Rhat)
**Solutions**:
```{r troubleshoot-convergence}
# Increase iterations
result <- fit_model(model_spec, data, iter = 4000)

# Increase adapt_delta
result <- fit_model(model_spec, data, control = list(adapt_delta = 0.99))

# Check for data issues
summary(trial_data)  # Look for outliers, missing values
```

**Problem**: Slow fitting
**Solutions**:
```{r troubleshoot-speed}
# Use more cores (if available)
result <- fit_model(model_spec, data, cores = 4)

# Start with smaller dataset
small_data <- trial_data[1:50, ]
result <- fit_model(model_spec, small_data)
```

### Prior Specification Issues

**Problem**: Understanding prior conversion
```{r troubleshoot-priors}
# Check how priors are converted
priors <- PriorSpecification()
priors <- add_prior(priors, "sigma", "half_cauchy", NA, 5, 0)
brms_priors <- to_brms_prior(priors)
print(brms_priors)  # Shows student_t(3, 0, 5)
```

## Next Steps

Once you have a working model:

1. **Experiment with different priors** - Try skeptical vs. optimistic priors
2. **Add more predictors** - Include additional covariates  
3. **Try different model families** - Binary outcomes (binomial), count data (Poisson)
4. **Conduct sensitivity analyses** - Compare results across different specifications
5. **Work with real data** - Apply to your actual clinical trial data

## Advanced Topics

For more advanced analyses, see additional vignettes:

- **Prior Specification Guide**: Detailed prior selection strategies
- **Model Diagnostics**: Advanced convergence checking
- **Sensitivity Analysis**: Comparing multiple model specifications  
- **Multi-trial Analysis**: Analyzing multiple studies together

## Getting Help

If you encounter issues:

1. Check this vignette first
2. Review function documentation: `?function_name`
3. Browse other vignettes: `browseVignettes("bayestrials")`
4. Report bugs: [GitHub Issues](https://github.com/AartikSarma/bayestrials/issues)

```{r session-info}
sessionInfo()
```