---
title: "Prior Specification Guide"
author: "bayestrials Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prior Specification Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Prior Specification in bayestrials

This guide covers everything you need to know about specifying priors in bayestrials, from basic concepts to advanced techniques.

```{r setup}
library(bayestrials)
```

## Understanding Priors in bayestrials

### The PriorSpecification System

bayestrials uses a structured approach to prior specification that:

1. **Validates** prior specifications before model fitting
2. **Converts** priors to the appropriate brms format
3. **Documents** your prior choices for reproducibility
4. **Enables** easy sensitivity analyses

```{r basic-prior-system}
# Create a PriorSpecification object
priors <- PriorSpecification()
print(priors)

# Add priors one by one
priors <- add_prior(priors, "b_treatment", "normal", 0, 2.5)
priors <- add_prior(priors, "Intercept", "normal", 0, 10)

print(priors)
```

### Supported Distributions

bayestrials supports all major prior distributions:

```{r supported-distributions}
# Normal priors (most common)
priors <- PriorSpecification()
priors <- add_prior(priors, "b_treatment", "normal", location = 0, scale = 1.5)

# Student-t priors (robust to outliers)
priors <- add_prior(priors, "b_age", "student_t", location = 0, scale = 1, df = 3)

# Half-Cauchy priors (for scale parameters)
priors <- add_prior(priors, "sigma", "half_cauchy", location = NA, scale = 5, lb = 0)

# Gamma priors (for positive parameters)
priors <- add_prior(priors, "shape", "gamma", location = 2, scale = 0.5)

# Beta priors (for probabilities)
priors <- add_prior(priors, "p", "beta", location = 2, scale = 5)

print(priors)
```

## Prior Selection Strategies

### Weakly Informative Priors (Recommended)

Weakly informative priors provide gentle regularization without overwhelming the data:

```{r weakly-informative}
# For treatment effects (typically small to moderate)
treatment_prior <- PriorSpecification()
treatment_prior <- add_prior(treatment_prior, "b_treatment", "normal", 0, 2.5)

# For continuous predictors (standardized)
treatment_prior <- add_prior(treatment_prior, "b_age", "normal", 0, 1.0)

# For intercepts (based on outcome scale)
treatment_prior <- add_prior(treatment_prior, "Intercept", "normal", 50, 20)

# For residual SD (half-Cauchy with reasonable scale)
treatment_prior <- add_prior(treatment_prior, "sigma", "half_cauchy", NA, 5, 0)

print(treatment_prior)
```

### Informative Priors (Expert Knowledge)

When you have prior knowledge or strong clinical beliefs:

```{r informative-priors}
# Strong belief in treatment effect
informative_priors <- PriorSpecification()

# Expect positive treatment effect around 3 units
informative_priors <- add_prior(informative_priors, "b_treatment", "normal", 3, 1)

# Age effect based on literature
informative_priors <- add_prior(informative_priors, "b_age", "normal", -0.1, 0.05)

# Tight prior on intercept
informative_priors <- add_prior(informative_priors, "Intercept", "normal", 45, 5)

print(informative_priors)
```

### Skeptical Priors (Conservative Analysis)

For skeptical analyses or regulatory submissions:

```{r skeptical-priors}
# Skeptical about treatment effects
skeptical_priors <- PriorSpecification()

# Prior centered on null effect with tight variance
skeptical_priors <- add_prior(skeptical_priors, "b_treatment", "normal", 0, 0.5)

# Minimal effects for other predictors
skeptical_priors <- add_prior(skeptical_priors, "b_age", "normal", 0, 0.25)

# Standard intercept
skeptical_priors <- add_prior(skeptical_priors, "Intercept", "normal", 50, 10)

print(skeptical_priors)
```

## Parameter-Specific Guidelines

### Treatment Effects

```{r treatment-effect-priors}
# Binary treatment coding (0/1)
# Effect size interpretation depends on outcome scale

# Continuous outcomes (difference in means)
priors <- PriorSpecification()
priors <- add_prior(priors, "b_treatment", "normal", 0, 2.5)  # Allows ±5 unit effects

# Binary outcomes (log-odds)
priors <- add_prior(priors, "b_treatment", "normal", 0, 1.5)   # Allows ORs ~0.2 to 5

# Count outcomes (log-rate ratio)  
priors <- add_prior(priors, "b_treatment", "normal", 0, 1.0)   # Allows rate ratios ~0.4 to 2.7
```

### Covariate Effects

```{r covariate-priors}
# Standardized continuous covariates
priors <- PriorSpecification()

# Age, BMI, etc. (typically small effects)
priors <- add_prior(priors, "b_age_std", "normal", 0, 0.5)

# Binary covariates (sex, comorbidity)
priors <- add_prior(priors, "b_male", "normal", 0, 1.0)

# Categorical covariates (site, severity level)
priors <- add_prior(priors, "b_site2", "normal", 0, 1.5)
priors <- add_prior(priors, "b_site3", "normal", 0, 1.5)
```

### Scale Parameters

```{r scale-parameters}
# Residual standard deviation (continuous outcomes)
priors <- PriorSpecification()
priors <- add_prior(priors, "sigma", "half_cauchy", NA, 5, 0)

# Random effect standard deviations
priors <- add_prior(priors, "sd_site", "half_cauchy", NA, 2.5, 0)

# Overdispersion parameters (negative binomial)
priors <- add_prior(priors, "shape", "gamma", 0.01, 0.01)
```

## Prior Conversion System

### How bayestrials Handles Priors

bayestrials automatically converts your prior specifications to brms format:

```{r prior-conversion}
# Your specification
my_priors <- PriorSpecification()
my_priors <- add_prior(my_priors, "sigma", "half_cauchy", NA, 5, 0)

# What brms sees (automatic conversion)
brms_format <- to_brms_prior(my_priors)
print(brms_format)
# Shows: student_t(3, 0, 5) - equivalent to half-Cauchy!
```

### Common Conversions

```{r common-conversions}
# Half-Cauchy → Student-t (bounded)
half_cauchy <- add_prior(PriorSpecification(), "sigma", "half_cauchy", NA, 2.5, 0)
print(to_brms_prior(half_cauchy))

# Normal → Normal (unchanged)
normal_prior <- add_prior(PriorSpecification(), "b_x", "normal", 0, 1)
print(to_brms_prior(normal_prior))

# Student-t → Student-t (with df parameter)
t_prior <- add_prior(PriorSpecification(), "b_x", "student_t", 0, 1, df = 3)
print(to_brms_prior(t_prior))
```

## Sensitivity Analysis with Priors

### Comparing Prior Specifications

```{r sensitivity-setup}
# Generate example data
data <- generate_synthetic_data(n_observations = 100, seed = 123)

# Base model specification
model_spec <- create_model_spec(
  outcome = "outcome_continuous",
  predictors = "treatment + age_centered",
  family = "gaussian"
)
```

```{r sensitivity-priors}
# Define different prior sets
neutral_priors <- PriorSpecification()
neutral_priors <- add_prior(neutral_priors, "b_treatment", "normal", 0, 2.5)
neutral_priors <- add_prior(neutral_priors, "b_age_centered", "normal", 0, 1.0)
neutral_priors <- add_prior(neutral_priors, "Intercept", "normal", 10, 5)
neutral_priors <- add_prior(neutral_priors, "sigma", "half_cauchy", NA, 5, 0)

skeptical_priors <- PriorSpecification()  
skeptical_priors <- add_prior(skeptical_priors, "b_treatment", "normal", 0, 1.0)
skeptical_priors <- add_prior(skeptical_priors, "b_age_centered", "normal", 0, 0.5)
skeptical_priors <- add_prior(skeptical_priors, "Intercept", "normal", 10, 5)
skeptical_priors <- add_prior(skeptical_priors, "sigma", "half_cauchy", NA, 5, 0)

optimistic_priors <- PriorSpecification()
optimistic_priors <- add_prior(optimistic_priors, "b_treatment", "normal", 2, 1.5)
optimistic_priors <- add_prior(optimistic_priors, "b_age_centered", "normal", 0, 1.0) 
optimistic_priors <- add_prior(optimistic_priors, "Intercept", "normal", 10, 5)
optimistic_priors <- add_prior(optimistic_priors, "sigma", "half_cauchy", NA, 5, 0)
```

```{r sensitivity-fitting}
# Fit models with different priors
attr(model_spec, "priors") <- neutral_priors
neutral_model <- fit_model(model_spec, data, chains = 1, iter = 500)

attr(model_spec, "priors") <- skeptical_priors  
skeptical_model <- fit_model(model_spec, data, chains = 1, iter = 500)

attr(model_spec, "priors") <- optimistic_priors
optimistic_model <- fit_model(model_spec, data, chains = 1, iter = 500)
```

```{r sensitivity-comparison}
# Compare treatment effect estimates
neutral_est <- extract_estimates(neutral_model)["b_treatment", "Estimate"]
skeptical_est <- extract_estimates(skeptical_model)["b_treatment", "Estimate"] 
optimistic_est <- extract_estimates(optimistic_model)["b_treatment", "Estimate"]

print(paste("Neutral:", round(neutral_est, 2)))
print(paste("Skeptical:", round(skeptical_est, 2)))
print(paste("Optimistic:", round(optimistic_est, 2)))
```

## Best Practices

### 1. Start with Weakly Informative Priors

```{r best-practice-1}
# Good default approach
standard_priors <- PriorSpecification()
standard_priors <- add_prior(standard_priors, "b_treatment", "normal", 0, 2.5)
standard_priors <- add_prior(standard_priors, "Intercept", "normal", 0, 10)
standard_priors <- add_prior(standard_priors, "sigma", "half_cauchy", NA, 5, 0)
```

### 2. Document Your Prior Choices

```{r best-practice-2}
# Always document reasoning
# For treatment effect: Normal(0, 2.5) allows effects up to ±5 units
# which covers clinically meaningful range for this outcome

# For intercept: Normal(0, 10) is vague but prevents extreme values
# For sigma: Half-Cauchy(0, 5) is standard weakly informative choice
```

### 3. Check Prior Sensitivity

```{r best-practice-3}
# Always run sensitivity analyses
# Compare neutral vs. skeptical priors
# Ensure conclusions are robust to prior choice
```

### 4. Visualize Your Priors

```{r prior-visualization}
# Plot prior distributions to understand implications
plot_priors(standard_priors, model_spec)
```

## Advanced Prior Specification

### Hierarchical Priors

```{r hierarchical-priors}
# For random effects models
hierarchical_priors <- PriorSpecification()

# Fixed effects
hierarchical_priors <- add_prior(hierarchical_priors, "b_treatment", "normal", 0, 2.5)

# Random effect standard deviations  
hierarchical_priors <- add_prior(hierarchical_priors, "sd_site", "half_cauchy", NA, 2.5, 0)

# Correlations (LKJ prior)
hierarchical_priors <- add_prior(hierarchical_priors, "cor_site", "lkj", 2, NA)
```

### Custom Prior Functions

For complex priors not covered by standard distributions:

```{r custom-priors}
# Example: Mixture of normals for bimodal prior
# This would require custom Stan code - contact package maintainers
```

## Troubleshooting

### Common Prior Issues

**Problem**: Model won't fit with specific priors
```{r troubleshoot-priors}
# Check prior conversion
problematic_priors <- PriorSpecification()
problematic_priors <- add_prior(problematic_priors, "b_x", "normal", 0, 0.001)  # Too tight!

# Diagnosis
brms_version <- to_brms_prior(problematic_priors)
print(brms_version)  # Check if reasonable

# Solution: Use more reasonable scale
fixed_priors <- PriorSpecification()
fixed_priors <- add_prior(fixed_priors, "b_x", "normal", 0, 1.0)  # Better!
```

**Problem**: Prior-data conflict warnings
```{r prior-data-conflict}
# Highly informative prior conflicts with data
# Either:
# 1. Relax the prior
# 2. Investigate data quality  
# 3. Accept the conflict and proceed with caution
```

## Summary

Key takeaways for prior specification in bayestrials:

1. **Start simple**: Use weakly informative priors
2. **Document choices**: Explain your reasoning
3. **Check sensitivity**: Compare different prior sets
4. **Trust the conversion**: bayestrials handles brms formatting
5. **Visualize**: Plot priors to understand their implications

```{r session-info}
sessionInfo()
```