---
title: "Troubleshooting Guide for bayestrials"
author: "bayestrials Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting Guide for bayestrials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Comprehensive Troubleshooting Guide

This guide provides solutions to common problems encountered when using bayestrials, organized by category with step-by-step solutions.

```{r setup}
library(bayestrials)
```

## Installation and Setup Issues

### Problem: Package Installation Fails

**Error Messages:**
- "compilation terminated"
- "tools for package building"  
- "unable to install dependencies"

**Solution:**

```{r install-troubleshoot}
# Step 1: Check R version (must be 4.0+)
R.version.string

# Step 2: Update R if needed
# Download from https://cran.r-project.org/

# Step 3: Install/update core dependencies first
dependencies <- c("rstan", "brms", "rstantools", "StanHeaders")
for(pkg in dependencies) {
  if(!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
  }
}

# Step 4: Try bayestrials installation again
devtools::install_github("AartikSarma/bayestrials")

# Step 5: Verify installation
library(bayestrials)
packageVersion("bayestrials")
```

### Problem: Stan/brms Compilation Fails

**Platform-Specific Solutions:**

**macOS:**
```bash
# Terminal commands
xcode-select --install
# Follow installation prompts, then restart R

# In R:
```

```{r macos-stan}
# Configure compiler
writeLines('CXX14FLAGS=-O3 -march=native -mtune=native', 
           file = file.path(Sys.getenv("HOME"), ".R", "Makevars"))

# Test compilation
library(rstan)
example(stan_model, package = "rstan", run.dontrun = TRUE)
```

**Windows:**
```{r windows-stan}
# Check Rtools installation
Sys.which("make")
Sys.which("gcc")

# If empty, download and install Rtools
# https://cran.r-project.org/bin/windows/Rtools/
# Ensure "Add rtools to system PATH" is checked

# Configure for Windows
file.edit(file.path(Sys.getenv("HOME"), ".Renviron"))
# Add: PATH="${RTOOLS40_HOME}\usr\bin;${PATH}"

# Restart R and test
library(rstan)
```

**Linux:**
```bash
# Ubuntu/Debian
sudo apt update
sudo apt install build-essential r-base-dev libv8-dev

# CentOS/RHEL
sudo yum groupinstall "Development Tools"  
sudo yum install R-devel v8-devel
```

### Problem: Memory Issues During Installation

```{r memory-install}
# Increase memory limits
options(repos = c(CRAN = "https://cloud.r-project.org"))
Sys.setenv("_R_CHECK_FORCE_SUGGESTS_" = "FALSE")

# Install in clean session
if(!requireNamespace("pkgbuild", quietly = TRUE)) {
  install.packages("pkgbuild")
}

# Build with specific memory settings
withr::with_envvar(c("_R_CHECK_FORCE_SUGGESTS_" = "FALSE"), {
  devtools::install_github("AartikSarma/bayestrials")
})
```

## Model Fitting Problems

### Problem: Model Won't Converge

**Symptoms:**
- Rhat > 1.01
- Low effective sample size
- "chains haven't mixed" warnings

**Systematic Diagnosis:**

```{r convergence-diagnosis}
# Step 1: Check your model and data
print(model_spec)
str(data)
summary(data)

# Step 2: Look for obvious issues
# - Extremely large/small values
# - Missing data patterns
# - Collinearity between predictors
cor(data[, c("age", "baseline_score", "treatment")])

# Step 3: Start with simplest model
simple_spec <- create_model_spec(
  outcome = "outcome_continuous",
  predictors = "treatment",  # Just treatment
  family = "gaussian"
)

# Simple priors
simple_priors <- PriorSpecification()
simple_priors <- add_prior(simple_priors, "b_treatment", "normal", 0, 2.5)
simple_priors <- add_prior(simple_priors, "Intercept", "normal", 10, 5)
simple_priors <- add_prior(simple_priors, "sigma", "half_cauchy", NA, 5, 0)

attr(simple_spec, "priors") <- simple_priors
```

**Progressive Solutions:**

```{r convergence-solutions}
# Solution 1: More iterations
result <- fit_model(simple_spec, data, 
                   iter = 4000, warmup = 2000, chains = 4)

# Solution 2: Better adaptation  
result <- fit_model(simple_spec, data,
                   control = list(adapt_delta = 0.95, max_treedepth = 12))

# Solution 3: Different initialization
result <- fit_model(simple_spec, data,
                   init = "random",  # or custom init values
                   init_r = 0.5)

# Solution 4: Reparameterization (for complex models)
# Center continuous predictors
data$age_centered <- scale(data$age, center = TRUE, scale = FALSE)[,1]
data$baseline_centered <- scale(data$baseline_score, center = TRUE, scale = FALSE)[,1]

# Update model
centered_spec <- create_model_spec(
  outcome = "outcome_continuous", 
  predictors = "treatment + age_centered + baseline_centered"
)
```

### Problem: Divergent Transitions

**Understanding Divergences:**
Divergent transitions indicate the Hamiltonian Monte Carlo sampler encountered numerical problems, usually due to:
1. Highly curved posterior geometry
2. Poor parameterization  
3. Inappropriate priors
4. Data scaling issues

**Solutions:**

```{r divergent-solutions}
# Solution 1: Increase adapt_delta
result <- fit_model(model_spec, data,
                   control = list(adapt_delta = 0.99))

# Solution 2: Check prior reasonableness
# Look at data scale first
summary(data$outcome_continuous)  # e.g., range 5-15

# Adjust priors to data scale
reasonable_priors <- PriorSpecification()
reasonable_priors <- add_prior(reasonable_priors, "Intercept", "normal", 10, 3)  # Center on data mean
reasonable_priors <- add_prior(reasonable_priors, "b_treatment", "normal", 0, 2)   # Reasonable effect size
reasonable_priors <- add_prior(reasonable_priors, "sigma", "half_cauchy", NA, 2, 0)  # Match data variability

# Solution 3: Standardize predictors
data$age_std <- scale(data$age)[,1]
data$baseline_std <- scale(data$baseline_score)[,1]

standardized_spec <- create_model_spec(
  outcome = "outcome_continuous",
  predictors = "treatment + age_std + baseline_std"
)

# Solution 4: Use robust family if outliers suspected
robust_spec <- create_model_spec(
  outcome = "outcome_continuous",
  predictors = "treatment + age_std + baseline_std", 
  family = "student_t"  # Heavy-tailed distribution
)
```

### Problem: Extremely Slow Fitting

```{r slow-fitting}
# Profile the problem
system.time({
  test_result <- fit_model(model_spec, data[1:50, ],  # Small subset
                          chains = 1, iter = 100)
})

# Solutions for speed:

# 1. Parallel processing
result <- fit_model(model_spec, data,
                   cores = parallel::detectCores() - 1,
                   chains = 4)

# 2. Reduce complexity temporarily
# Test with smaller dataset
small_data <- data[sample(nrow(data), 200), ]

# 3. Simplify model first
basic_spec <- create_model_spec(
  outcome = "outcome_continuous",
  predictors = "treatment"  # Add complexity gradually
)

# 4. Consider variational inference for exploration
vi_result <- fit_model(model_spec, data,
                      algorithm = "meanfield",  # Faster, less accurate
                      iter = 1000)

# 5. Optimize Stan settings
result <- fit_model(model_spec, data,
                   control = list(
                     adapt_engaged = TRUE,
                     adapt_delta = 0.8,      # Don't start too high
                     max_treedepth = 10      # Default value
                   ))
```

## Prior Specification Issues

### Problem: Don't Know What Priors to Use

```{r prior-guidance}
# Step 1: Understand your outcome scale
summary(data$outcome_continuous)
hist(data$outcome_continuous)

# Step 2: Use weakly informative priors as starting point
priors <- PriorSpecification()

# For treatment effect - allow reasonable range
# If outcome ranges 0-20, treatment effect might be ±5 units max
priors <- add_prior(priors, "b_treatment", "normal", 0, 2.5)

# For intercept - center near data mean with wide range
outcome_mean <- mean(data$outcome_continuous, na.rm = TRUE)
outcome_sd <- sd(data$outcome_continuous, na.rm = TRUE)
priors <- add_prior(priors, "Intercept", "normal", outcome_mean, outcome_sd * 2)

# For residual SD - based on outcome variability
priors <- add_prior(priors, "sigma", "half_cauchy", NA, outcome_sd, 0)

# For standardized covariates
priors <- add_prior(priors, "b_age_std", "normal", 0, 1)  # Unit normal for standardized

print(priors)

# Step 3: Visualize prior implications
plot_priors(priors, model_spec)
```

### Problem: Prior-Data Conflict

```{r prior-data-conflict}
# Diagnose prior-data conflict
# Run prior predictive check
prior_check <- brm(
  build_formula(model_spec), 
  data = data,
  prior = to_brms_prior(priors),
  sample_prior = "only",
  chains = 2, iter = 500
)

# Check if prior predictions make sense
prior_predictions <- predict(prior_check)
summary(prior_predictions)

# Compare to actual data
summary(data$outcome_continuous)

# If mismatch, adjust priors:
# Example: Prior predicted range -100 to 200, but data is 5 to 15

adjusted_priors <- PriorSpecification()
adjusted_priors <- add_prior(adjusted_priors, "Intercept", "normal", 10, 3)  # Narrower
adjusted_priors <- add_prior(adjusted_priors, "b_treatment", "normal", 0, 1.5)  # More conservative
adjusted_priors <- add_prior(adjusted_priors, "sigma", "half_cauchy", NA, 2, 0)  # Smaller scale
```

### Problem: Understanding Prior Conversion

```{r prior-conversion-help}
# Check what happens to your priors
original_priors <- PriorSpecification()
original_priors <- add_prior(original_priors, "sigma", "half_cauchy", NA, 5, 0)
original_priors <- add_prior(original_priors, "b_treatment", "normal", 0, 2.5)

# See brms conversion
brms_version <- to_brms_prior(original_priors)
print(brms_version)

# Understanding the conversion:
# half_cauchy(0, 5) becomes student_t(3, 0, 5) 
# This is mathematically equivalent for positive values

# Verify equivalence by sampling
set.seed(123)
half_cauchy_samples <- abs(rt(1000, df = 1)) * 5  # Half of Cauchy(0, 5)
student_t_samples <- rt(1000, df = 3) * 5         # Student-t(3, 0, 5) 

# Both should have similar distributions for positive values
summary(half_cauchy_samples)
summary(pmax(0, student_t_samples))  # Truncate at 0
```

## Data-Related Issues

### Problem: Missing Data

```{r missing-data-handling}
# Assess missingness pattern
library(VIM)
VIM::aggr(data, col = c('navyblue','red'), numbers = TRUE, sortVars = TRUE)

# Check if data is missing completely at random (MCAR)
library(naniar)
naniar::vis_miss(data)

# Solutions:

# 1. Complete case analysis (if <5% missing)
complete_data <- na.omit(data)
cat("Original:", nrow(data), "Complete:", nrow(complete_data), 
    "Lost:", nrow(data) - nrow(complete_data), "\n")

# 2. Multiple imputation (if >5% missing)
library(mice)
imputed <- mice(data, m = 5, method = 'pmm', printFlag = FALSE)
completed_data <- complete(imputed)

# 3. Model-based handling (brms can handle some missingness)
# Missing predictors: brms will exclude those observations
# Missing outcomes: automatic exclusion with warning

# Check what brms will do with your data
model_result <- fit_model(model_spec, data)  # Will show warnings about excluded rows
```

### Problem: Outliers and Influential Observations

```{r outlier-handling}
# Identify potential outliers
# 1. Visual inspection
par(mfrow = c(2, 2))
hist(data$outcome_continuous, main = "Outcome Distribution")
boxplot(data$outcome_continuous, main = "Outcome Boxplot")
plot(data$age, data$outcome_continuous, main = "Age vs Outcome")
plot(data$treatment, data$outcome_continuous, main = "Treatment vs Outcome")

# 2. Statistical identification
# Z-score method (values > 3 SD from mean)
outcome_z <- abs(scale(data$outcome_continuous))
outlier_rows <- which(outcome_z > 3)
cat("Potential outliers:", length(outlier_rows), "out of", nrow(data), "\n")

# 3. Robust analysis options:

# Option A: Remove outliers (with justification)
clean_data <- data[-outlier_rows, ]

# Option B: Use robust error distribution
robust_spec <- create_model_spec(
  outcome = "outcome_continuous",
  predictors = "treatment + age + baseline_score",
  family = "student_t"  # Heavy-tailed, robust to outliers
)

# Option C: Transform outcome
# Log transform (for positive outcomes)
if(all(data$outcome_continuous > 0, na.rm = TRUE)) {
  data$log_outcome <- log(data$outcome_continuous)
  
  log_spec <- create_model_spec(
    outcome = "log_outcome", 
    predictors = "treatment + age + baseline_score"
  )
}

# Option D: Winsorize extreme values
library(DescTools)
data$outcome_winsorized <- Winsorize(data$outcome_continuous, probs = c(0.05, 0.95))
```

### Problem: Collinearity Between Predictors

```{r collinearity-handling}
# Detect collinearity
# Correlation matrix
cor_matrix <- cor(data[, c("age", "baseline_score", "biomarker_level")], 
                 use = "complete.obs")
print(cor_matrix)

# Variance Inflation Factors (VIF)
# First fit a simple linear model
lm_model <- lm(outcome_continuous ~ treatment + age + baseline_score + biomarker_level, 
               data = data)
car::vif(lm_model)  # VIF > 5 indicates problems

# Solutions:

# 1. Remove highly correlated predictors
# Keep only one from highly correlated pair
reduced_spec <- create_model_spec(
  outcome = "outcome_continuous",
  predictors = "treatment + age + baseline_score"  # Remove biomarker if cor > 0.8 with baseline
)

# 2. Principal component analysis
# Create PCA of correlated predictors
pca_vars <- data[, c("age", "baseline_score", "biomarker_level")]
pca_result <- prcomp(pca_vars, scale = TRUE)
data$PC1 <- pca_result$x[, 1]
data$PC2 <- pca_result$x[, 2]

pca_spec <- create_model_spec(
  outcome = "outcome_continuous", 
  predictors = "treatment + PC1 + PC2"
)

# 3. Regularization through priors
# Use tighter priors on correlated predictors
regularized_priors <- PriorSpecification()
regularized_priors <- add_prior(regularized_priors, "b_treatment", "normal", 0, 2.5)
regularized_priors <- add_prior(regularized_priors, "b_age", "normal", 0, 0.5)          # Tighter
regularized_priors <- add_prior(regularized_priors, "b_baseline_score", "normal", 0, 0.5) # Tighter  
regularized_priors <- add_prior(regularized_priors, "Intercept", "normal", 10, 5)
regularized_priors <- add_prior(regularized_priors, "sigma", "half_cauchy", NA, 3, 0)
```

## Interpretation Issues

### Problem: Understanding Bayesian Output

```{r interpretation-help}
# Extract and understand results
result <- fit_model(model_spec, data, chains = 2, iter = 1000)
estimates <- extract_estimates(result)

# Key columns explained:
# Estimate: Posterior mean (best point estimate)
# Est.Error: Posterior standard deviation (uncertainty)
# Q2.5, Q97.5: 95% credible interval bounds
# Rhat: Convergence diagnostic (should be ≤ 1.01)
# Bulk_ESS, Tail_ESS: Effective sample sizes (should be > 400)

print(estimates)

# Focus on treatment effect
treatment_est <- estimates["b_treatment", ]
cat("Treatment effect estimate:", round(treatment_est$Estimate, 3), "\n")
cat("95% Credible interval: [", round(treatment_est$Q2.5, 3), ", ", 
    round(treatment_est$Q97.5, 3), "]\n", sep = "")

# Probability calculations
posterior_samples <- extract_posterior(result, "b_treatment")
prob_positive <- mean(posterior_samples > 0)
prob_meaningful <- mean(posterior_samples > 1.0)  # Adjust threshold as needed

cat("Probability treatment effect > 0:", round(prob_positive * 100, 1), "%\n")
cat("Probability treatment effect > 1.0:", round(prob_meaningful * 100, 1), "%\n")
```

### Problem: Model Comparison and Selection

```{r model-comparison-help}
# Fit multiple models for comparison
model1 <- fit_model(create_model_spec("outcome_continuous", "treatment"), data)
model2 <- fit_model(create_model_spec("outcome_continuous", "treatment + age"), data)  
model3 <- fit_model(create_model_spec("outcome_continuous", "treatment + age + baseline_score"), data)

# Compare using information criteria
comparison <- compare_models(list(
  "treatment_only" = model1,
  "treatment_age" = model2,
  "full_model" = model3
))

print(comparison)

# Interpret comparison:
# WAIC: Widely Applicable Information Criterion (lower is better)  
# LOO: Leave-One-Out Cross-Validation (lower is better)
# SE: Standard error of the difference
# ELPD_diff: Difference in expected log predictive density

# Model averaging if models are close
if(abs(comparison$waic_diff[2]) < 2 * comparison$se_diff[2]) {
  cat("Models have similar predictive performance - consider model averaging\n")
} else {
  best_model <- rownames(comparison)[which.min(comparison$waic)]
  cat("Best model:", best_model, "\n")
}
```

## Getting Additional Help

### Diagnostic Checklist

Before asking for help, run through this checklist:

```{r diagnostic-checklist}
# System information
cat("=== System Information ===\n")
sessionInfo()
cat("\nbayestrials version:", packageVersion("bayestrials"), "\n")

# Data information  
cat("\n=== Data Summary ===\n")
str(data)
cat("Missing values:", sum(is.na(data)), "\n")
cat("Outcome range:", range(data$outcome_continuous, na.rm = TRUE), "\n")

# Model information
cat("\n=== Model Specification ===\n")
print(model_spec)
validate_model_spec(model_spec)

# Prior information
if(!is.null(attr(model_spec, "priors"))) {
  cat("\n=== Prior Specifications ===\n")
  print(attr(model_spec, "priors"))
  
  cat("\n=== brms Prior Conversion ===\n")
  print(to_brms_prior(attr(model_spec, "priors")))
}

# Fitting diagnostics (if model was fitted)
if(exists("result") && !is.null(result)) {
  cat("\n=== Model Diagnostics ===\n")
  diagnostics <- check_diagnostics(result)
  print(diagnostics)
}
```

### Preparing a Minimal Reproducible Example

```{r minimal-example}
# Create minimal example for bug reports
set.seed(123)
minimal_data <- data.frame(
  treatment = factor(rep(c("A", "B"), each = 25)),
  age = rnorm(50, 65, 10),
  outcome = rnorm(50, 10, 3)
)

minimal_spec <- create_model_spec(
  outcome = "outcome",
  predictors = "treatment + age", 
  family = "gaussian"
)

minimal_priors <- PriorSpecification()
minimal_priors <- add_prior(minimal_priors, "b_treatmentB", "normal", 0, 2)
minimal_priors <- add_prior(minimal_priors, "b_age", "normal", 0, 1)
minimal_priors <- add_prior(minimal_priors, "Intercept", "normal", 10, 5)
minimal_priors <- add_prior(minimal_priors, "sigma", "half_cauchy", NA, 3, 0)

attr(minimal_spec, "priors") <- minimal_priors

# This should reproduce your issue:
# minimal_result <- fit_model(minimal_spec, minimal_data)
```

### Resources for Further Help

1. **Package Documentation**: `?bayestrials` and function help pages
2. **Vignettes**: `browseVignettes("bayestrials")`
3. **brms Documentation**: https://paul-buerkner.github.io/brms/
4. **Stan Documentation**: https://mc-stan.org/users/documentation/
5. **GitHub Issues**: https://github.com/AartikSarma/bayestrials/issues

```{r session-info}
sessionInfo()
```